### 2012-09-04 sections/implementation/execution-workflow.tex
* back to 8 pages   * reduced all figure sizes to 0.9/linewidth   * removed tcpdump, wireshark, libpcap paragraph with citations
* revision with professor's comments   * introduction     * updated title, added vlad's IM 2011 paper     * reduced from bibtex entries, back to page 8     * added collado:2009 and sperotto:2010 references   * implementation     * flow query intermediate format     * grouper, grouper aggregation     * merger (TODO)

A custom C library has been written to directly read/write data in the
\texttt{flow-tools} format. The library sequentially reads the complete
flow-records into memory to support random access required for relative
filtering. Each flow-record is stored in a \texttt{char} array and the offsets
to each field are stored in separate \texttt{structs} as shown in listing
\ref{lst:c-lib-struct}. The array of such records are indexed allowing fast
retrieval in $O(1)$ time.

@@

\subsubsection{Splitter} \ac{NFQL} uses identifiers to reference a flow record
in the \texttt{char} array. This gives away the need to every time copy all
the flow-records when moving ahead in the processing pipeline. As a result,
there is no dedicated splitter stage in the execution engine. Each branch
references the flow records from a common memory location. This helps keep the
memory costs at a minimum when multiple branches are involved.

\subsubsection{Filter}
\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=1.0\linewidth]{filter-fv1-fv2}
    \caption{Filter Stage: NFQL vs NFQL (optimized)}
    \label{fig:fv1-fv2-filter}
  \end{center}
\end{figure}

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to fill in the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \ac{NFQL} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publically available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of the optimized
\ac{NFQL} implementation against the one defined by the flow query language as
shown in Fig. \ref{fig:fv1-fv2-filter}. The filter stage implementation with
these optimizations runs 10 times faster and is more pronounced on higher
ratios.

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in \ac{NFQL}
is to sort the filtered recordset on all the requested grouping terms of a
clause. This helps the execution engine perform a nested binary search to
reduce the linear pass to a fairly small filtered recordset. However, the
number of elements upon which the search has to be performed needs to be known
at each level of the binary lookup. In order to avoid a linear run to count
the number of elements, the parent level binary search invocation returns a
boundary with the first and last element to enable such a calculation. This
helps the grouper perform faster search lookups to find records that must
group together in $O(n*lg(k))$ time with a preprocessing step taking
$O(p*n*lg(n))$ in the average case, where $n$ is the number of filtered
records, $p$ is the number of grouping terms in a clause and $k$ is the number
of unique filtered records.

\begin{figure}[h!]
  \begin{center}
    \includegraphics*[width=1.0\linewidth]{grouper-fv2-bsearch-fv2}
    \caption{Grouper Stage: NFQL (Generic) vs NFQL (Specific)}
    \label{fig:fv1-fv2-grouper}
  \end{center}
\end{figure}

The grouping approach has further been optimized when the filtered records are
grouped for equality. In such a scenario, the need to search for unique
records and a  subsequent binary search goes away.  The groups can now be
formed in $O(n)$ time with the same preprocessing step taking $O(p*n*lg(n)$.
The performance evaluation of the grouper handling this special case against
its behavior when handling generic cases is shown in Fig.
\ref{fig:fv1-fv2-grouper}. Profiling the performance of the grouper on queries
that produce higher ratios reveal that a significant amount of time is taken
in binary search which is eliminated in the special case of equality
comparisons.

The resultant group records are a conglomeration of multiple flow records with
some common characteristics. Some of the non-common characteristics can also
be aggregated into a single value using group aggregations as defined in the
query. Such an aggregated group record is again mapped to a NetFlow $v5$
record template. This allows the aggregated group records to be written to
file as a representative of all its members.

@@

The merger is used to relate groups from different branches according to a
merging criterean. The implementation is not trivial since the number of
branches that need to be spawned is read from the query and is not known until
runtime. As a result, an iterator that can provide all possible permutations
of $m-$tuple (where $m$ is the number of branches) group record $ID$s was
needed.  The result of the iterator can then be used to make a match. The
merger stage, begins by initializing this iterator passing it the number of
branches, and information about each branch. Then, it loops over to get a new
$m-$tuple of group record $ID$s on each iteration until the iterator returns
\texttt{false}.

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records however can be reduced by sorting the group records on the field
offsets used for a match. The \ac{NFQL} implementation optimizes the merger to
skip over iterator permutations when a state of a current field offset value
may not allow any further match beyond the index in the current branch.  For
such an optimization to work, the filtered group records must be sorted in the
order of field offsets specified in the merger clause. Specifying the filtered
group records in any other order may lead to undefined behavior. This means,
that if the same field offsets were used in the grouper stage, the terms in
the group clause can be rearranged by the query designer to align with the
order of terms in the merger clause.

>>>>

A custom C library has been written to directly read/write data stored in
\texttt{flow-tools} format. The library sequentially reads the flow-records
into memory to support random access required for relative filtering. Each
flow-record is stored in a \texttt{char} array and the offsets to each field
are stored in separate \texttt{structs} as shown in listing
\ref{lst:c-lib-struct}. The array of such records are indexed allowing fast
retrieval in $O(1)$ time.

@@

\subsubsection{Splitter} \ac{NFQL} uses identifiers to reference a flow record
in the \texttt{char} array. This eliminated the need to copy all the
flow-records when moving ahead in the processing pipeline. As a result, there
is no dedicated splitter stage in the execution engine. Each branch references
the flow records from a common memory location. This helps keep the memory
costs at a minimum when multiple branches are involved.

\subsubsection{Filter}
\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=0.9\linewidth]{filter-fv1-fv2}
    \caption{Filter Stage: NFQL vs NFQL (optimized)}
    \label{fig:fv1-fv2-filter}
  \end{center}
\end{figure}

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \ac{NFQL} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. The filtered records are saved in a common location from
where they are referenced by each branch. This helps keep the memory costs at
a minimum when multiple branches are involved.  We used the publically
available Trace 7 from the SimpleWeb \cite{simpleweb} to compare the
performance of the optimized \ac{NFQL} implementation against the one defined
by the flow query language as shown in Fig.  \ref{fig:fv1-fv2-filter}. The
filter stage implementation with these optimizations runs 10 times faster and
is more pronounced on higher ratios.

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in \ac{NFQL}
is to sort the filtered recordset on the field offsets all the requested
grouping terms of a clause. This helps the execution engine perform a nested
binary search to reduce the linear pass to a fairly small filtered recordset.
However, the number of elements upon which the search has to be performed
needs to be known at each level of the binary lookup. In order to avoid a
linear run to count the number of elements, the parent level binary search
invocation returns a boundary with the first and last element to enable such a
calculation. This helps the grouper perform faster search lookups to find
records that must group together in $O(n*lg(k))$ time with a preprocessing
step taking $O(p*n*lg(n))$ in the average case, where $n$ is the number of
filtered records, $p$ is the number of grouping terms in a clause and $k$ is
the number of unique filtered records.

\begin{figure}[h!]
  \begin{center}
    \includegraphics*[width=0.9\linewidth]{grouper-fv2-bsearch-fv2}
    \caption{Grouper Stage: NFQL (Generic) vs NFQL (Specific)}
    \label{fig:fv1-fv2-grouper}
  \end{center}
\end{figure}

The grouping approach has further been optimized when the filtered records are
grouped for equality. In such a scenario, the need to search for unique
records and a  subsequent binary search goes away.  The groups can now be
formed in $O(n)$ time with the same preprocessing step taking $O(p*n*lg(n))$.
The performance evaluation of the grouper handling this special case against
its behavior when handling generic cases is shown in Fig.
\ref{fig:fv1-fv2-grouper}. Profiling the performance of the grouper on queries
that produce higher ratios reveal that a significant amount of time is taken
in binary search which is eliminated in the special case of equality
comparisons.

The resultant group records are a conglomeration of multiple flow records with
some common characteristics. Some of the non-common characteristics can also
be aggregated into a single value using group aggregations as defined in the
query. Such an aggregated group record is again mapped to a NetFlow $v5$
record template. This allows the aggregated group records to be written to a
file as a representative of all its members.

@@

The merger is used to relate groups from different branches according to a
merging criterean. The implementation is not trivial since the number of
branches that need to be spawned is read from the query and is not known until
runtime. As a result, an iterator that can provide all possible permutations
of $m-$tuple (where $m$ is the number of branches) group record $ID$s was
needed.  The result of the iterator can then be used to make a match. The
merger stage begins by initializing this iterator passing it the number of
branches, and information about each branch. Then, it loops over to get a new
$m-$tuple of group record $ID$s on each iteration until the iterator returns
\texttt{false}.

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. The \ac{NFQL} implementation optimizes the
merger to skip over iterator permutations when a state of a current field
offset value may not allow any further match beyond the index in the current
branch.  For such an optimization to work, the filtered group records must be
sorted in the order of field offsets specified in the merger clause.
Specifying the filtered group records in any other order may lead to undefined
behavior. This means, that if the same field offsets were used in the grouper
stage, the terms in the group clause can be rearranged by the query designer
to align with the order of terms in the merger clause.

### 2012-09-04 sections/implementation/flow-query-intermediate-format.tex
* revision with professor's comments   * introduction     * updated title, added vlad's IM 2011 paper     * reduced from bibtex entries, back to page 8     * added collado:2009 and sperotto:2010 references   * implementation     * flow query intermediate format     * grouper, grouper aggregation     * merger (TODO)

The entire JSON flow query is a collection of \ac{DNF} expressions. A \ac{DNF}
expression is a disjunction of conjunctive clauses. The elements of the
conjuctive clauses are terms. The clauses in the \ac{DNF} are OR'd together,
while the terms in each clause themselves are AND'd. The branchsets and each
\ac{DNF} expression of the pipeline stage is a \texttt{JSON array}.
\texttt{json-c} \cite{jsonc} is used to parse the flow query file. The
pipeline stages of the flow query language are also encapsulated in Python
classes and scripts have been written to facilitate the front-end parser to
serialize these pipeline objects to a \texttt{JSON} flow query.  The mapping
of the query to the structs defined in the execution engine is
not trivial. When reading the \texttt{JSON} query at runtime, the field
offsets of the NetFlow $v5$ record struct are read in as strings.  Utility
functions are defined that map the read field names to struct offsets and
field types and the operations to unique enum members.

@@

The \texttt{JSON} query can also trigger and disable the stages at runtime.
This means that one only has to supply the constructs that one wishes to use.
The constructs that are not defined in the \texttt{JSON} query are inferred by
the execution engine as a disable request. The execution engine uses disable
flags that are turned on when the query is parsed. These flags are used
throughout the engine to only enable the requested functionality.

>>>>

Each stage of the processing pipeline is expressed in the \texttt{JSON} query
as a \ac{DNF} expression. A \ac{DNF} expression is a disjunction of
conjunctive clauses. The elements of the conjuctive clauses are terms. The
clauses in the \ac{DNF} are OR'd together, while the terms in each clause
themselves are AND'd. \texttt{json-c} \cite{jsonc} is used to parse the flow
query file. The pipeline stages of the flow query language are also
encapsulated in Python classes and scripts have been written to facilitate the
front-end parser to serialize these pipeline objects to a \texttt{JSON} flow
query.  The mapping of the query to the structs defined in the execution
engine is not trivial. When reading the \texttt{JSON} query at runtime, the
field names of a NetFlow $v5$ record are read in as strings.  Utility
functions are defined that map the field names to internal struct offsets and
the field types and the operations to internal unique enum members.

@@

The \texttt{JSON} query can also disable the stages at runtime.  This means
that one only has to supply the constructs that one wishes to use.  The
constructs that are not defined in the \texttt{JSON} query are inferred by the
execution engine as a disable request. The execution engine uses disable flags
that are turned on when the query is parsed. These flags are used throughout
the engine to only enable the requested functionality.

### 2012-09-04 sections/introduction.tex
* back to 8 pages   * reduced all figure sizes to 0.9/linewidth   * removed tcpdump, wireshark, libpcap paragraph with citations
* revision with professor's comments   * introduction     * updated title, added vlad's IM 2011 paper     * reduced from bibtex entries, back to page 8     * added collado:2009 and sperotto:2010 references   * implementation     * flow query intermediate format     * grouper, grouper aggregation     * merger (TODO)

Flow capture today, has emerged out to be one of the favored network
measurement techniques. In this technique, packets traversing a monitoring
point are not captured raw, instead they are aggregated together based on some
common characteristics. The common characteristics are learnt by inspecting
the packet headers as they cross the monitoring point. Flow-records resulting
from such an aggregation are then exported to a collector for further
analysis. This not only reduces the amount of traffic at the monitoring point,
but also provides fine-grained control over the network data which was not
previously possible using SNMP interface-level queries.

NetFlow and \ac{IPFIX} are the two popular standards of IP flow information
export. NetFlow \cite{rfc3954} is a proprietary network protocol designed by
Cisco Systems. It allow routers to generate and export flow records to a
designated collector. NetFlow v$9$ provides flexibility of user-tailored
export templates, \ac{MPLS} and IPv$6$ support and a larger set of flow keys.
\ac{IPFIX} \cite{rfc5101} on the other hand is an open standard by IETF deemed
to be the logical successor of NetFlow v$9$ on which it is based. The novelty
of the standard lies in its ability to describe record formats at runtime
using templates based on an extensible and well-defined information model. The
data transfer mechanism is also simplistic and extensible by being
unidirectional and transport protocol agnostic.

@@

It goes to show that understanding these intricate traffic patterns require
sophisticated flow analysis tools that can mine flow records for such a usage.
Unfortunately current tools fail to deliver owing to their poor language
design and simplistic filtering methods.  We recently proposed a flow query
language design \cite{vmarinov:thesis:2009} that aims to cater to such needs.
It can process flow records, aggregate them into groups, apply absolute (or
relative) filters and invoke Allen interval algebra rules \cite{fallen:1983}.

In this paper, we introduce \ac{NFQL}, an efficient C implementation of the
flow query language. It is the next iteration of the first prototype
implementation, Flowy \cite{kkanev:2010}, which was written in Python.
\ac{NFQL} however is not just a reimplementation in a new language, but
the inner workings have been reimagined to allow the execution engine to
scale to real-world sized traces. This has been possible by replacing the
performance hit stages of the pipeline with crispier algorithms that lower
down the execution times to factor making them comparable to contemporary
flow analysis tools.

The rest of the paper is organized as follows. In section \ref{sec:design} we
describe the flow query language by discussing each stage of the processing
pipeline. In section \ref{sec:implementation} we introduce the inner workings
of \ac{NFQL}. It begins by providing an overview of the \ac{NFQL} components
and the structure of the intermediate format used to exchange messages. The
workflow of the execution engine is described next and is supplemented by a
number of performance optimizations made to make the implementation scale. It
is followed by performance evaluation results by comparing them against
contemporary flow processing tools in section \ref{sec:evaluation} and
concludes with a discussion on its current limitation and future outlook in
section \ref{sec:conclusion}.

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\linewidth]{nfql-pipeline}
\caption{NFQL Processing Pipeline \cite{vmarinov:2009}}
\label{fig:nfql-pipeline}
\end{figure*}

>>>>

Capturing network flow records has emerged out to be one of the favored
network measurement techniques. In this technique, packets traversing an
obvervation point are not captured raw, instead they are aggregated together
based on some common characteristics. The common characteristics are learnt by
inspecting the packet headers as they cross the observation point.
Flow-records resulting from such an aggregation are then exported to a
collector for further analysis. This reduces the amount of traffic aggregated
at the collector.

NetFlow and \ac{IPFIX} are the two popular protocols for IP flow information
export. NetFlow \cite{rfc3954} is a proprietary network protocol designed by
Cisco Systems. It allows routers to generate and export flow records to a
designated collector. NetFlow v$9$ provides flexibility of user-tailored
export templates, \ac{MPLS} and IPv$6$ support and a larger set of flow keys.
\ac{IPFIX} \cite{rfc5101} on the other hand is an open standard by IETF and
deemed to be the logical successor of NetFlow v$9$ on which it is based. The
novelty of the standard lies in its ability to describe record formats at
runtime using templates based on an extensible and well-defined information
model. The data transfer mechanism is also simplistic and extensible by being
unidirectional and transport protocol agnostic.

@@

Understanding intricate traffic patterns require sophisticated flow analysis
tools that can mine flow records for such a usage.  Unfortunately current
tools fail to deliver owing to their language design and simplistic filtering
methods.  We recently proposed a flow query language design
\cite{vmarinov:2009} that aims to cater to such needs.  It can process
flow records, aggregate them into groups, apply absolute (or relative) filters
and invoke Allen interval algebra rules \cite{fallen:1983}. The applicability
of the language can be seen from \cite{vperelman:2011} where the authors
formulate flow queries to identify flow signatures of popular applications.

In this paper, we introduce \ac{NFQL}, an efficient C implementation of the
flow query language. It is the next iteration of the first prototype
implementation, Flowy \cite{kkanev:2010}, which was written in Python.
\ac{NFQL}, however, is not just a reimplementation in a new language, but the
inner workings have been reimagined to allow the execution engine to scale to
real-world sized traces. This has been possible by replacing the performance
critical stages of the pipeline with crispier algorithms that lower the
execution times, making \ac{NFQL} comparable to contemporary flow analysis
tools.

The paper is organized as follows. In section \ref{sec:relatedwork} we survey
the current state-of-the-art flow-processing tools and reason how \ac{NFQL} is
different from each one of them. In section \ref{sec:design} we describe the
flow query language by discussing each stage of the processing pipeline. In
section \ref{sec:implementation} we introduce the inner workings of \ac{NFQL}.
It begins by providing an overview of the \ac{NFQL} architecture and the
structure of the intermediate format used to exchange messages. The workflow
of the execution engine is described next and is supplemented by a number of
performance optimizations made to make the implementation scale. The section
is followed by performance evaluations comparing \ac{NFQL} against
contemporary flow processing tools alongwith with a discussion on its current
limitation and future outlook in section \ref{sec:evaluation}, with conclusion
in section \ref{sec:conclusion}.

\begin{figure*}[!t]
\centering
\includegraphics[width=0.9\linewidth]{nfql-pipeline}
\caption{NFQL Processing Pipeline \cite{vmarinov:2009}}
\label{fig:nfql-pipeline}
\end{figure*}
